Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-21 paper.tex
Update storage size discussion for new benchmarks

Fig~\ref{fig-data-storage} shows compression performance 
on up to a million samples for chromosome 21, with 
the size of the genotype-matrix encoded as 1-bit per haploid
call included for reference.
% 2bit  1689.153258 G
% vcf   81.375831 G
% = 20.75 X, 4.8%
Gzip compressed VCF performs remarkably well, compressing 
the data to around 5\% of the 
minimal binary encoding of a biallelic genotype matrix
for 1 million samples. 
BCF provides a significant improvement in compression
performance over VCF (note the log-log scale). Genozip has 
superb compression, having far smaller file sizes that the 
other methods (although somewhat losing its advantage at 
larger sample sizes). Zarr and Savvy have 
very similar compression performance in this example (23.75GiB
and 21.25GiB for $10^6$ samples, respectively). When using
a chunk size of $10^4$ variants $\times$ $10^3$ samples, Zarr
storage is reduced to 22.07GiB (not shown).
It is remarkable that the simple approach of compressing
two dimensional chunks of the genotype matrix 
using the Zstandard compressor~\citep{collet2021rfc} and the 
bit-shuffle filter from Blosc~\cite{alted2010modern} 
(see Methods for details) produces 
compression levels competitive with the highly specialised methods
used by Savvy.

@@

The data shown in Fig~\ref{fig-whole-matrix-compute} for Zarr and Savvy
is based on custom programs written using their respective APIs
to implement the \texttt{af-dist} operation. The Zarr program uses
the Zarr-Python package to iterate over the decoded chunks of the 
genotype matrix and classifies genotypes within a chunk using a 14 line Python
function, accelerated using the Numba JIT compiler~\cite{lam2015numba}.
The allele frequencies and genotype counts are then analysed to produce 
the final counts within the allele frequency bins with 9 lines of 
Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
short and simple Python program is substantially faster than the 
equivalent compiled C using \texttt{htslib} APIs on BCF (6.9 hours
vs 20.6 hours for 1 million samples). 
% num_samples  num_sites      tool  user_time  sys_time     wall_time  total_time
%     1000000    7254858     savvy    8371.07      4.16   8377.718136  2.326453
%     1000000    7254858  bcftools   73939.32     46.09  74023.083678  20.551503
%     1000000    7254858      zarr   24709.64     29.32  24750.704171  6.871933
The fastest method is the 
C++ program written using the Savvy API. This would largely seem
to be due to Savvy's excellent genotype decoding performance
% zarr 1.2 GiB
% savvy 6.6 GiB
% zarr_nshf 2.7 GiB
% zarr_lz4_nshf 2.9 GiB
(up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
Fig~\ref{fig-whole-matrix-decode}).
% sav / zarr ratio: [0.68777206 0.75870945 0.69865406 0.64644888 0.68131752
% 0.89460906]
%     sequence_length  num_samples  num_sites                tool         size
% 58         48129895      1000000    7254858                zarr    23.752762
% 50         48129895      1000000    7254858         10kx1k.zarr    22.068925
% 51         48129895      1000000    7254858                 bcf    51.749294
% 52         48129895      1000000    7254858             genozip    10.691384
% 53         48129895      1000000    7254858  lz4.noshuffle.zarr   214.868005
% 54         48129895      1000000    7254858      noshuffle.zarr    31.208149
% 55         48129895      1000000    7254858                 sav    21.249436
% 56         48129895      1000000    7254858                 tsk     1.802636
% 57         48129895      1000000    7254858                 vcf    81.375831
% 59         48129895      1000000    7254858            zarr.zip    24.240338
% 5          48129895      1000000    7254858                2bit  1689.153258
[TODO revise this text!!]
Turning off the BitShuffle filter for the Zarr dataset,
however, leads to a substantial increase in decoding speed
(2.9GiB/s) at the cost of a roughly 25\% increase in storage
space (29.9GiB up from 22.1GiB for 1 million samples; data not
shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
decoding speed, but at the cost of much larger storage requirements 
(219GiB; data not shown).
Given the relatively small contribution of genotypes to the
overall storage of real datasets (see the Genomics England example)
and the frequency that they are likely to be accessed, using 
Zstandard without BitShuffle 
would seem like a good tradeoff in many cases.
This ability to easily tune compression performance
and decoding speed on a field-by-field basis is a major strong
point of Zarr. The \texttt{vcf2zarr} utility also provides
functionality to aid with such storage schema tuning.

A substantial advantage of Zarr is the availabilty of several 
independent implementations across different programming languages
(see Methods). Figure~\ref{fig-whole-matrix-compute-zarr-versions}
compares the CPU time required to perform the computations
of Figure~\ref{fig-whole-matrix-compute} 
for af-dist implementations using the Zarr-Python~\cite{zarrpython},
JZarr~\cite{jzarr},
and TensorStore~\citep{tensorstore} (C++ and Python) libraries.
The version using the Tensorstore C++ API is fastest, requiring 
about 20\% less time than Zarr-Python (which is used in other
benchmarks).

>>>>

Fig~\ref{fig-data-storage} shows compression performance 
on up to a million samples for chromosome 21, with 
the size of the genotype-matrix encoded as 1-bit per haploid
call included for reference.
% 2bit  1689.153258 G
% vcf   81.375831 G
% = 20.75 X, 4.8%
Gzip compressed VCF performs remarkably well, compressing 
the data to around 5\% of the 
minimal binary encoding of a biallelic genotype matrix
for 1 million samples. 
BCF provides a significant improvement in compression
performance over VCF (note the log-log scale). Genozip has 
superb compression, having far smaller file sizes that the 
other methods (although somewhat losing its advantage at 
larger sample sizes). Zarr and Savvy have 
very similar compression performance in this example (23.75GiB
and 21.25GiB for $10^6$ samples, respectively). When using
a chunk size of $10^4$ variants $\times$ $10^3$ samples, Zarr
storage is reduced to 22.07GiB for $10^6$ samples (data not shown;
see Figs~\ref{fig-compression-chunksize} and \ref{fig-compression-chunksize-finegrained}
for more analysis on the effect of chunk size on compression).
It is remarkable that the simple approach of compressing
two dimensional chunks of the genotype matrix 
using the Zstandard compressor~\citep{collet2021rfc} and the 
bit-shuffle filter from Blosc~\cite{alted2010modern} 
(see Methods for details) produces 
compression levels competitive with the highly specialised methods
used by Savvy.

@@

The data shown in Fig~\ref{fig-whole-matrix-compute} for Zarr and Savvy
is based on custom programs written using their respective APIs
to implement the \texttt{af-dist} operation. The Zarr program uses
the Zarr-Python package to iterate over the decoded chunks of the 
genotype matrix and classifies genotypes within a chunk using a 14 line Python
function, accelerated using the Numba JIT compiler~\cite{lam2015numba}.
The allele frequencies and genotype counts are then analysed to produce 
the final counts within the allele frequency bins with 9 lines of 
Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
short and simple Python program is substantially faster than the 
equivalent compiled C using \texttt{htslib} APIs on BCF (6.9 hours
vs 20.6 hours for 1 million samples). 
% num_samples  num_sites      tool  user_time  sys_time     wall_time  total_time
%     1000000    7254858     savvy    8371.07      4.16   8377.718136  2.326453
%     1000000    7254858  bcftools   73939.32     46.09  74023.083678  20.551503
%     1000000    7254858      zarr   24709.64     29.32  24750.704171  6.871933
The fastest method is the 
C++ program written using the Savvy API. This would largely seem
to be due to Savvy's excellent genotype decoding performance
% zarr 1.2 GiB
% savvy 6.6 GiB
% zarr_nshf 2.7 GiB
% zarr_lz4_nshf 2.9 GiB
(up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
Fig~\ref{fig-whole-matrix-decode}).
% sav / zarr ratio: [0.68777206 0.75870945 0.69865406 0.64644888 0.68131752
% 0.89460906]
%     sequence_length  num_samples  num_sites                tool         size
% 58         48129895      1000000    7254858                zarr    23.752762
% 50         48129895      1000000    7254858         10kx1k.zarr    22.068925
% 51         48129895      1000000    7254858                 bcf    51.749294
% 52         48129895      1000000    7254858             genozip    10.691384
% 53         48129895      1000000    7254858  lz4.noshuffle.zarr   214.868005
% 54         48129895      1000000    7254858      noshuffle.zarr    31.208149
% 55         48129895      1000000    7254858                 sav    21.249436
% 56         48129895      1000000    7254858                 tsk     1.802636
% 57         48129895      1000000    7254858                 vcf    81.375831
% 59         48129895      1000000    7254858            zarr.zip    24.240338
% 5          48129895      1000000    7254858                2bit  1689.153258
Turning off the BitShuffle filter for the Zarr dataset,
however, leads to a substantial increase in decoding speed
(2.7GiB/s) at the cost of a roughly 25\% increase in storage
space (31.2GiB up from 23.75GiB for 1 million samples; data not
shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
decoding speed (2.9GiB/s), but at the cost of much larger storage requirements 
(214GiB; data not shown).
Given the relatively small contribution of genotypes to the
overall storage of real datasets (see the Genomics England example)
and the frequency that they are likely to be accessed, using 
Zstandard without BitShuffle 
would seem like a good tradeoff in many cases.
This ability to easily tune compression performance
and decoding speed on a field-by-field basis is a major strong
point of Zarr. The \texttt{vcf2zarr} utility also provides
functionality to aid with such storage schema tuning.

Another advantage of Zarr is the availabilty of several 
independent implementations across different programming languages
(see Methods). Figure~\ref{fig-whole-matrix-compute-zarr-versions}
compares the CPU time required to perform the computations
of Figure~\ref{fig-whole-matrix-compute} 
for af-dist implementations using the Zarr-Python~\cite{zarrpython},
JZarr~\cite{jzarr},
and TensorStore~\citep{tensorstore} (C++ and Python) libraries.
The version using the Tensorstore C++ API is fastest, requiring 
about 20\% less time than Zarr-Python (which is used in other
benchmarks).
